---
title: 'Getting Started with R'
output: pdf_document 
#bibliography: references.bib
urlcolor: red 
---
\setcounter{table}{0}
\renewcommand{\thetable}{A\arabic{table}}
\renewcommand{\thefigure}{A\arabic{figure}}

```{r setup, include=FALSE}
library(knitr)
library(formatR)
knitr::opts_chunk$set(echo = TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=55))

```

# How are R and R Studio Related? 

\begin{itemize}
\item R is a statistical programming language that runs data and statistical analysis. 
\item RStudio is a powerful user interface to R with many convenient features that makes the use of R easier. 
\item We will use R with RStudio, but there are other user interfaces that some people use. 
\item Unless you have prior programming experience (or have a preference for another text editor), it is recommended that you use RStudio.
\item At its simplest, you can think the difference between R and R Studio with the help of a useful analogy: R is like a car’s engine while RStudio is like a car’s dashboard!
\begin{figure} [h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=.9\linewidth]{analogy.png}\\
\label{fig1}
\end{tabular} \\
\end{center}
\caption{The Difference Between R and R Studio, \textit{Source: Modern Dive}} 
\end{figure}
\end{itemize}

# Installing R and R Studio 

\begin{itemize}
\item We need to install R \textit{before} installing RStudio. 
\item Please make sure to select the installer appropriate for your operating system. 
\end{itemize}

## Video Tutorials

\begin{itemize}
\item Chris Bail at Duke University has an instructive video that shows how to install R and RStudio step by step. You can find Bail's video \href{https://www.youtube.com/watch?v=ulIv0NiVTs4&feature=emb_logo}{here}. 
\end{itemize}

\begin{itemize}
\item If you have difficulty in following Bail's video tutorial for installation, you can find Rutgers Data Librarian Ryan Womack’s video useful, which goes at a slower pace. You can find Womack's video 
\href{https://www.youtube.com/watch?v=5rOR31bmoL8}{here}.

\end{itemize}

\clearpage

## Written Tutorial

\begin{itemize}
\item If you instead want a written tutorial for installation the following links will be helpful: 
\begin{itemize}
\item \textbf{For installing R:} Go to \href{https://cran.r-project.org}{Cran} and follow the instructions for installing R.
\item \textbf{For installing RStudio:} Go to \href{https://rstudio.com/products/rstudio/download/}    {RStudio} and follow the instructions for installing RStudio.
\end{itemize}
\end{itemize}

# Using R via RStudio: Opening R Studio 

\begin{itemize}
\item After installing R and R Studio, you’ll have two new programs/applications you can open on your computer.
\item To work in R, we’ll \textit{always} open RStudio. 
\item When you first open RStudio, you will see something similar to Figure A2 with three main windows: 
\begin{figure} [h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=.8\linewidth]{rstudiointerface1.png}\\
\label{fig2}
\end{tabular}
\end{center}
\caption{RStudio interface to R, three main windows}
\end{figure}

\clearpage 

\item Once you open files, such as R scripts, a text editor window will also open in the top left. 
\item If you only see three main windows, let's open a new R script file, which will activate the text editor window. To do this click File $\rightarrow$ New File $\rightarrow$ R Script. 
\item You should see something similar to Figure A3 with four main windows: 
\end{itemize}

\begin{figure} [h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=.8\linewidth]{rstudioface2.png}\\
\label{fig3}
\end{tabular}
\end{center}
\caption{RStudio interface to R, opening the Text Editor Window/R Script, four main windows}
\end{figure}

# How do I begin using R?

\begin{itemize}
\item Statistical software programs like Excel, or SPSS provide \textit{point-and-click interfaces}. \item Unlike these programs, R is an \textit{interpreted language}. This means that we have to type in commands written in R code.
\item We will explore these in detail in the workshop, but you can do some quick tryouts to get familiar with the structure of R.
\end{itemize}

## Console Window (bottom left window)

\begin{itemize}
\item Console window is the bottom left window in your RStudio. 
\item You can think this window as your scratch paper. Here, you can do some quick computations by using R as a calculator or try out some commands.  

\clearpage

\item Let's do some quick trials. Type the following in your console and and hit <enter> or <return>: 
\item The console will show the output of any commands you type.
\end{itemize}

```{r echo=T, results='hide'}
5 + 82
```

```{r echo=T, results='hide'}
5 + 82
```

```{r echo=T, results='hide'}
5 - 3 
```

```{r echo=T, results='hide'}
50*4 
```

```{r echo=T, results='hide'}
10^2
```

\begin{itemize}
\item Your screen should look like Figure A4: 
\begin{figure} [h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=1\linewidth]{console.png}\\
\label{fig4}
\end{tabular}
\end{center}
\caption{RStudio interface to R, Console}
\end{figure}
\end{itemize}

## Text Editor Window (upper left window)

\begin{itemize}
\item Once we open an R script ﬁle in RStudio, the text editor will appear as one of the windows. It can then be used to write our code.
\item Text editor window is the main place where we will tell R what we want to do. 
\item You can think this as your main notebook. When we are conducting an analysis in R, we will save our R Script and work inside of it. 
\item Let's do some quick trials. Type the following in your text editor window: 
\end{itemize}

```{r echo=T, results='hide'}
2 + 2
```

```{r echo=T, results='hide'}
sum22 <- 2 + 2
```

\begin{itemize}
\item To run a command on your R script: 
\begin{itemize}
\item Place your cursor at the end of the line of code (far right), and hit <command + return> on a Mac; <control + return> on Windows. 
\item Or, place your cursor at the end of the line of code (far right), and click `Run' in the top right corner of the upper-left window. \footnote{Source: Katie McCabe's MLE book}
\end{itemize}
\item After you execute the code, you will see the result in your console for the first code. You will see the second code in your console and it will also pop up in your environment. We will discuss why this happens in the workshop. 
\item Your screen should look like Figure A5: 
\end{itemize}

\begin{figure} [h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=1\linewidth]{editor.png}\\
\label{fig5}
\end{tabular}
\end{center}
\caption{RStudio interface to R, Text Editor/R Script}
\end{figure}

\clearpage

## Environment/History Window (upper right window)

\begin{itemize}
\item The upper right lists R objects and a history of executed R commands.
\end{itemize}

## Files/Plots/Help/Viewer Window (bottom right window)

\begin{itemize}
\item The bottom right will show multiple things: (1) files that are in your working directory, (2) graphs that you create, (3) help documents, (4) other features.
\end{itemize}

\begin{itemize}
\item[] We will revise these steps and continue with working in and preparing an R script in the workshop. Feel free to send an email and/or post a message on Slack if you have any questions. 

\item[] Hope to see you soon! 
\end{itemize}
