---
title: 'Introduction to R, Day 3 Handout'
output: pdf_document 
#bibliography: references.bib
urlcolor: red 
---
\setcounter{table}{0}
\renewcommand{\thetable}{A\arabic{table}}
\renewcommand{\thefigure}{A\arabic{figure}}

```{r setup, include=FALSE}
library(foreign)
library(knitr)
library(formatR)
knitr::opts_chunk$set(echo = TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=55))
```

In this handout, we cover the following topics and R commands: 

## Topics 

- Correlation
- Linear Regression
- Visualizing Bivariate Relationships
- More on Visualization: Visualization with ggplot2

## R Commands 

- Calculating correlation using `cor()`
- Fitting linear regression models using `lm()`
- Summarizing linear regression models using `summary()`
- Extracting coefﬁcients, standard errors, t-values, and p-values from `lm`
- Obtaining model coefficients using `coef()`
- Creating a scatterplot  using `plot()` 
- Adding best fit lines to scatter plots using `abline()`
- Placing multiple plots in one ﬁgure using `par`
- Obtaining residuals using `resid()`
- Obtaining fitted values using `fitted()`
- Creating residual plots using `plot()`
- Creating graphs with `ggplot2()`

\clearpage

# Correlation

- Correlation is one of the most frequently used statistics to summarize the relationship between two variables, which measures the degree to which two variables are associated to each other. \footnote{Some people refer to as correlation coefficient.}

- More specifically, the `correlation coefficient` tells us whether the change in the average values of one variable are related to the change in the average values of the other variable.

- Correlations ranges from -1 to 1. A score of -1 means a perfect negative association and a score of 1 a perfect positive asssociation between the two variables, i.e. 0.81 would count as a strong positive correlation. A score of zero means that there is no association.

\begin{figure} [h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=.8\linewidth]{correlation.png}\\
\end{tabular}
\end{center}
\caption{Source: Imai, QSS, 142}
\end{figure}

\clearpage

- We calculate correlations in R with the `cor()` function. It takes two arguments x and y, as follows: 

`cor(x,y)`

- To makes things more concrete, we will illustrate how to calculate a correlation coefficient in R with some running examples. 

## Trade Union Membership and Government Structure

- In their 1991 exchange article, Wallterstein and Stephens discusses the relationship between 
union membership and government structure (the control of government by left-wing and socialist parties).  \footnote{Example adapted from Kosuke Imai's online teaching materials}

- The table below shows the variables and their description: 

------------------------------------------------------------------------------
- `country`. Country
- `union`. Percentage of workers who belong to a union
- `left`. Extent to which parties of the left have controlled government
- `size`. Size of the labor force
- `concent` Measure of economic concentration in top four industries
------------------------------------------------------------------------------

- Load the data 
```{r}
tradedat <- read.csv("day3data/tradeunion.csv")
```

- Calculate the correlation using the `cor()` function
```{r}
cor(tradedat$union, tradedat$left)
## we can round to two decimal places 
round(cor(tradedat$union, tradedat$left), digits = 2) 
round(cor(tradedat$union, tradedat$left), 2) 
```

- This correlation coefficient suggests that there is a strong positive relationship between union membership and control of government by left-wing parties. 

\clearpage

# Correlation and Scatterplots 

- Scatterplots allows us to visually compare two variables measured on the same set of units by plotting the value of one variable against that of the other for each unit.

- We can visually inspect the correlation between union membership and control of government by left-wing parties. 

- To create a scatterplot in R, we use the `plot()` function. 

- The syntax for this function is 

`plot(x, y)` 

\textit{where x and y are vectors of horizontal and vertical coordinates, respectively.}

- Similar to yesterday, let's take a two step procedure to plotting:

 1. Creating a very simple plot
 2. Iteratively making our graph beautiful  

### Step 1: A Very Simple Plot

```{r , echo=F, fig.height=6, fig.width=6}
plot(tradedat$left, tradedat$union)
```

\clearpage

### Step 2: Iteratively Making it Better

- Label the main title and the x-axis, decrease the font size of the title

```{r , echo=F, fig.height=4, fig.width=4}
plot(tradedat$left, tradedat$union,
     xlab = "Left Government Control", ylab = "Union Labor Force Percentage",
     main = "Union Rates and Party Control of Government", cex.main = 0.7, cex.lab=0.7)
```

- Fix the x-axis and y-axis limits, set the size of points, 

```{r , echo=F, fig.height=4, fig.width=4}
plot(tradedat$left, tradedat$union,cex = 1, xlim = c(0, 120), ylim = c(0, 120),
     xlab = "Left Government Control", ylab = "Union Labor Force Percentage",
     main = "Union Rates and Party Control of Government", cex.main = 0.7, cex.lab = 0.7)
```

\clearpage 

- Add a 45 degree line 

```{r , echo=F, fig.height=4, fig.width=4}
plot(tradedat$left, tradedat$union,cex = 1, xlim = c(0, 120), ylim = c(0, 120),
     xlab = "Left Government Control", ylab = "Union Labor Force Percentage",
     main = "Union Rates and Party Control of Government", cex.main = 0.7, cex.lab = 0.7)
abline(0, 1, lty = 2, col = "red") 
```

\clearpage 

# Linear Regression 

\noindent \textit{The linear regression model estimated with ordinary least squares (OLS) is a workhorse model in Political Science. Even when a scholar uses a more advanced method that may make more accurate assumptions about his or her data— such as probit regression, a count model, or even a uniquely crafted Bayesian model— the researcher often draws from the basic form of a model that is linear in the parameters. By a similar token, many of the R commands for these more advanced techniques use functional syntax that resembles the code for estimating a linear regression. Therefore, an understanding of how to use R to estimate, interpret, and diagnose the properties of a linear model lends itself to sophisticated use of models with a similar structure (Monogan III, Political Analysis Using R)}

- To ﬁt a linear regression model in R, we use the `lm()` (`l`linear `m`model) function. This function takes a formula of the form Y ~ X as the main argument that connects one variable y to one (or more) variables x, taken from a specified data frame. 

`lm(y ~ x, data)`

- y is the outcome (dependent) variable and x is a single independent variable. In R, the intercept will be automatically added to the regression model.

- The argument we feed into the `lm` function is of the form `y ~ x`, where `~` shows that it is a formula.  In other words, we are assuming a linear model that has the following form: 
$$
y_i = \alpha + \beta x_i + \epsilon_i
$$

\textit{Note:The lm function allows to estimate the coefficients $\widehat \alpha$ and 
$\widehat \beta$ such that the line (best fit line) is as close to the data as possible.}

## Trade Union Membership and Leftist Government Control Revisited 

- To explore linear regression and its diagnostics, we will continue from the study that focuses on the relationship between trade union membership and government structure. 

- Let's `regress` trade union membership on the control of government by left-wing parties. 

```{r}
lm1 <- lm(union ~ left, data = tradedat)
lm1 
```

- `lm1` is the object that contains all the information we need to interpret the linear regression. 

- The output shows that the estimated intercept is 39.8841 whereas the estimated slope is 0.6604. The slope represents the average change in Y given a one unit change in X. In other words, according to the model, the estimated effect of a one percentage point increase in the control of government by left parties on the trade union membership is captured by the estimated regression coefficient `0.3764`. 

- To access `the coefﬁcients` from the regression, we can call them directly within `lm1` object. 

```{r}
lm1$coef # Both coefficients
lm1$coef[1] # The intercept
lm1$coef[2] # The slope
```

- When needed, we can also store information about our coefficients by creating an object: 

```{r}
coefs <- lm1$coef
coefs[1] 
coefs[2] 
```

- Is the effect we observe significant? Now, we can explore the components of the regression model in more detail.  To obtain a detailed summary of the regression output, we can use the `summary()` function.

```{r}
summary(lm1)
```

- The `p-value`, in the last column, is approximately 0.001, which is below the standard 0.05 threshold. Therefore, we can reject the null hypothesis that there is no relationship between trade union membership and control of government by left-wing parties. However, we didn't consider many other factors that might have an effect on the trade union membership. 

- We can also use `coef` here to access the coefﬁcients, standard errors, t-statistics, and p-values of the regression:

```{r}
summary(lm1)$coef
```

```{r}
summary(lm1)$coef
```

- Suppose that we want to access the p-value for the effect of leftist government control

```{r}
summary(lm1)$coef[2, 4]
```

- Lastly, let's calculate the confidence intervals for th effect of leftist government control
```{r}
mean.gov <- summary(lm1)$coef[2, 1] # mean of leftist gov
se.gov <- summary(lm1)$coef[2, 2] # standard error of leftist gov 
ci.gov <- c(mean.gov - 1.96 * se.gov, mean.gov + 1.96 * se.gov)
ci.gov
```

- Alternatively, we can calculate the confidence interval in a more straightforward way: 
```{r}
confint(lm1,level = 0.95)
```

## Regression and Scatterplots 

- Previously, we computed the correlation between trade union membership and leftist government control. Using a scatterplot, we had created the following scatterplot: 

```{r , echo=F, fig.height=4, fig.width=4}
plot(tradedat$left, tradedat$union,cex = 1, xlim = c(0, 120), ylim = c(0, 120),
     xlab = "Left Government Control", ylab = "Union Labor Force Percentage",
     main = "Union Rates and Party Control of Government", cex.main = 0.7, cex.lab = 0.7)
abline(0, 1, lty = 2, col = "red") 
```

-  Based on our regression model, we can update this scatterplot by adding a `regression line` to the graph. To do so , we will use the `abline()` function: 

```{r , echo=F, fig.height=4, fig.width=4}
plot(tradedat$left, tradedat$union,cex = 0.4, pch = 5, xlim = c(0, 120), ylim = c(0, 120),
     xlab = "Left Government Control", ylab = "Union Labor Force Percentage",
     main = "Union Rates and Party Control of Government", cex.main = 0.7, cex.lab = 0.7)
abline(0, 1, lty = 2, col = "red") 
abline(lm1, lty = 2, col = "purple") 
```

- Let's put these two scatterplots side by side in one figure. To do so, we will use a function called `par`.

> par(mfrow = c(number of rows, number of columns))

\textit{where the par function set the value of mfrow which determines how our plots will be arranged in the plots window}

- In our case, we only have two scatterplots. It would make sense to put them side by side by setting our number of rows to `1` and the number of columns to `2`: 

```{r}
par(mfrow = c(1, 2))
plot(tradedat$left, tradedat$union,cex = 1, xlim = c(0, 120), ylim = c(0, 120),
     xlab = "Left Government Control", ylab = "Union Labor Force Percentage",
     main = "Union Rates and Party Control of Government", cex.main = 0.7)
abline(0, 1, lty = 2, col = "red") 
plot(tradedat$left, tradedat$union,cex = 1, xlim = c(0, 120), ylim = c(0, 120),
     xlab = "Left Government Control", ylab = "Union Labor Force Percentage",
     main = "Union Rates and Party Control of Government", cex.main = 0.7, cex.lab = 0.7)
abline(0, 1, lty = 2, col = "red") 
abline(lm1, lty = 2, col = "purple") 
```

\clearpage

## Obtaining Residuals 

- The function `resid()` extracts the vector of residuals from a linear regression. 

```{r}
resid.lm1 <- resid(lm1)
resid.lm1
```

## Obtaining Fitted Values 

- The function `fitted()` obtains the fitted values from a linear regression. 

```{r}
fitted.lm1 <- fitted(lm1)
fitted.lm1
```

```{r} 
par(mfcol = c(1, 2))
plot(tradedat$left, resid(lm1), xlab = "Leftist Government Control",
ylab = "Residuals", main = "Residual Plot", cex.main = 0.8)
abline(h = 0, col = "red", lty = 2, lwd = 2) # Adds a zero line
plot(fitted(lm1), resid(lm1), xlab = "Fitted Values", ylab = "Residuals",
main = "Fitted Values vs. Residuals", cex.main = 0.8)
abline(h = 0, col = "red", lty = 2, lwd = 2)
```

\clearpage

# More on Visualization: Visualization with ggplot2

**Kieran Healy, DViz, Chapter**

## `Base R graphics` vs. `ggplot2`

- Previously, we highlighted that in R there are a lot of ways to the same thing. This is true especially with visualization. 

- R comes with `built-in functionality` for data analysis and visualization, typicall referred as `base R`. In the context of visualization, we have used the `base R` to create density plots, histograms, and scatterplots. 

- However, there are `R packages` that extend R's functionalities and capabilities.

- When it comes to data visualization, `ggplot2` created by \href{https://ggplot2-book.org/}
{Hadley Wickham} by far the most popular one. 

- You might likely hear a debate over using `base R graphics` vs. `ggplot2`. Some people feel more comfortable using `base R graphics` that we have been using while others find `ggplot2` cooler and mor effective to visualize data. 

\begin{figure} [h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=.8\linewidth]{basevsgg.jpg}\\
\end{tabular}
\end{center}
\caption{Source: Barbara Fusinska}
\end{figure}

-  If you want to know more, you might find \href{https://flowingdata.com/2016/03/22/comparing-ggplot2-and-r-base-graphics/}{this article} that discusses the differences between `base R graphics` and `ggplot2`. 

- We will learn both. And, it's really up to you to decide to use only one of them or both! 

\clearpage

## Installing packages

- We need to install a package only once, but we will need to load it at the beginning of each R session with `library()` if you want to use the tools it contains.

- To install a package: 
> install.packages("packagename")

- To load the package 
> library("packagename")

- The `tidyverse` is a set of related packages for R. The `ggplot2` package for data visualization is one of its components. You might also find \href{https://www.tidyverse.org/packages/} {other packages} of this master package such as `dplyr`, `tidyr` handy to do various things. `ggplot2` is a powerful, versatile, and widely used visualization package for R (Wickham 2016). 

- To cut a long story short, we will install the `tidyverse` package which includes but not limited to the `ggplot2` package for visualization. 

```{r, echo = F, results="hide" }
## Load packages
library(tidyverse) 
```

- To check your loaded packages, you can use `search()`
```{r}
## Load packages
search() 
```

## The Intuition Behind ggplot2 

- The most important thing to get used to with ggplot is the way you use it to think about the `logical structure` of your plot. Once you get used to the structure, you will realize that visualization process will more or less involve very similar sequence of steps. 

- Let's see step by step how `ggplot2` works with an example. 

- We will use a dataset called `gapminder` for our running example. This dataset includes variables about a large number of countries, each observed over several years.

- Our aim will be to plot `life expectancy` against `GDP per capita` for all country-years in the data. 
- This dataset belongs to an R package called `gapminder. That's why we also need to install and load a package called `gapminder`. 

```{r}
library(gapminder)
```

- Let's quickly obtain some information about this dataset before we visualize the relationship between life expectancy and GDP per capita. 

- Let's check the first few rows of the dataset quicky. 

- Check first few rows 
```{r}
head(gapminder)
```

- Number of observations and variables 
```{r}
dim(gapminder)
```

- Summary of each variable 
```{r}
summary(gapminder)
```

## Ggplot Step by Step 

- The first two arguments will help us to create the `core component` of our plot: `data` and `mapping`. 

- Once we complete these two main steps, the process will be `additive`: we will add `layers` to this core component by using `+` \footnote{In R, usually functions cannot simply be added to objects. Rather, they take objects as inputs and produce objects as outputs. But the objects created by ggplot() are special. This makes it easier to assemble plots one piece at a time, and to inspect how they look at every step}
}

1. Tell `ggplot()` what data we are using. The first argument we feed into the function is the  `data` argument. 

```{r}
myp <- ggplot(data = gapminder)
```

2. We need tell which specific variables we would like to `map` to each other. The second argument we give is the `mapping` argument. 

```{r}
myp <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp))
```

- Let's try to display to display the plot by printing our object 

```{r , echo=T, fig.height=3, fig.width=4}
myp
```

- What do you see?

- Thus far, we didn't passed any argument about what `sort` of plot we want. Do we want a scatterplot, histogram, trend line, boxplot etc.? We need to tell this. Some commonly used plot types are as follows: 

-----------------------------------------------------
- `geom_point()`. for scatter plots, dot plots, etc.
- `geom_boxplot()`. for boxplots!
- `geom_line()`. for trend lines, time series, etc.
-----------------------------------------------------

3. To tell the sort of plot we want, we need to `add` a new layer to the core component of our plot. Simply, adding a new layer means adding a new argument to the plot. Because we want to create a scatterplot, we will use the function `geom_point()` to add a layer. 

```{r , echo=T, fig.height=3, fig.width=4}
myp <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) + geom_point()
myp
```

3.1 We can change the sort of the plot, for example, we can plot a smoothed line shaded in a ribbon showing the standard error for the line. We will use `geom_smooth()` as our layer. 

```{r , echo=T, fig.height=3, fig.width=4}
myp <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) + geom_smooth()
myp
```

3.2 We can also see the points and lines together by adding both layers at once. 

```{r , echo=T, fig.height=3, fig.width=4}
myp <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +   geom_point() +
  geom_smooth() 
myp
```

3.3. In your R console, you will see a message that tells \textit{the geom_smooth() function is using a method called gam}, which means that it fit a `generalized additive model`. Let's change this to a `linear model` by setting the argument of the `geom_smooth()` function. 

```{r , echo=T, fig.height=3, fig.width=4}
myp <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +   geom_point() +
  geom_smooth(method = "lm") 
myp
```

4.GDP per capita is not normally distributed across country years. The graph will look better if we scale the x-axis to a log. We will add another layer called `scale_x_log10()`

```{r , echo=T, fig.height=3, fig.width=4}
myp <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +   geom_point() +
  geom_smooth(method = "lm") + scale_x_log10()
myp
```

5. It seems like we have a simple plot that looks okay. Let's polish this a little bit by adding a main title and a subtitle, fix the names of x-axis and y-axis lables, 

```{r , echo=T, fig.height=3, fig.width=4}
myp <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +   geom_point() +
  geom_smooth(method = "lm") + scale_x_log10() + 
  labs(x = "GDP Per Capita", y = "Life Expectancy in Years", title = "Economic Growth and Life Expectancy",
       subtitle = "Data points are country-years")
myp
```


6. Let's make the points transparent. To do so, we will feed an argument into the `geom_point` that takes care of the points. `alpha` is an aesthetic property that points (and some other plot elements) have, and to which variables can be mapped. It controls how transparent the object will appear when drawn.

```{r , echo=T, fig.height=3, fig.width=4}
myp <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +   geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") + scale_x_log10() + 
  labs(x = "GDP Per Capita", y = "Life Expectancy in Years", title = "Economic Growth and Life Expectancy",
       subtitle = "Data points are country-years")
myp
```

7. In the gapminder dataset, there is a variable called continent. Let's try whether we can color individual data points by continent. Because continent is a variable, we will feed this into the core `mapping()` argument. 

```{r , echo=T, fig.height=3, fig.width=4}
myp <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp, color = continent)) +   geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") + scale_x_log10() + 
  labs(x = "GDP Per Capita", y = "Life Expectancy in Years", title = "Economic Growth and Life Expectancy",
       subtitle = "Data points are country-years")
myp
```


8. We can also shade the standard error ribbon of each line to match its dominant color for each continent.

```{r , echo=T, fig.height=3, fig.width=4}
myp <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, 
                                              y = lifeExp, color = continent, fill = continent)) +   geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") + scale_x_log10() + 
  labs(x = "GDP Per Capita", y = "Life Expectancy in Years", title = "Economic Growth and Life Expectancy",
       subtitle = "Data points are country-years")
myp
```

8.1 Instead of having five seperate smooth lines, we might consider having one smooth line while keeping the colors of the points. To do so, we need to tell that we need to change the colors of the points. Therefore, we will use the layer that we added `geom_points()` instead of the `mapping()` function. 

```{r , echo=T, fig.height=3, fig.width=4}
myp <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(mapping = aes(colour=continent)) +
  geom_smooth(method = "lm") + scale_x_log10() + 
  labs(x = "GDP Per Capita", y = "Life Expectancy in Years", title = "Economic Growth and Life Expectancy",
       subtitle = "Data points are country-years")
myp
```

9 Let's also change the grey background color by adding another layer to our function, which sets the theme.  

```{r , echo=T, fig.height=3, fig.width=4}
myp <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(mapping = aes(colour=continent)) +
  geom_smooth(method = "lm") + scale_x_log10() + 
  labs(x = "GDP Per Capita", y = "Life Expectancy in Years", title = "Economic Growth and Life Expectancy",
       subtitle = "Data points are country-years") + theme_minimal()
myp
```

10. Lastly, let's save our first plot with ggplot! We will use the ``ggsave()` function. R will save the plot to the place where we had set our working directory. 

```{r , echo=F, fig.height=3, fig.width=4}
myp <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(mapping = aes(colour=continent)) +
  geom_smooth(method = "lm") + scale_x_log10() + 
  labs(x = "GDP Per Capita", y = "Life Expectancy in Years", title = "Economic Growth and Life Expectancy",
       subtitle = "Data points are country-years")
myp
ggsave("myfirstgg1.pdf") # save as a pdf
```

# $\mho$ Practice Exercises

1. In the morning session of the workshop, we created a scatterplot for the relationship between trade union membership and leftist government control. Let’s try to create the same scatterplot by using ggplot2!

```{r} 
plot(tradedat$left, tradedat$union,cex = 0.8, pch = 19, xlim = c(0, 120), ylim = c(0, 120),
     xlab = "Left Government Control", ylab = "Union Labor Force Percentage",
     main = "Union Rates and Party Control of Government
     with base R", cex.main = 1, cex.lab = 1)
abline(0, 1, lty = 1, col = "red") 
abline(lm1, lty = 1, col = "blue") 
```

```{r}
ggplot(data = tradedat, mapping = aes(x = left, y = union)) + geom_point() + 
  geom_smooth(method ="lm", se = FALSE) +
  geom_abline(aes(intercept = 0, slope = 1, col = "red")) + 
  xlim(0, 115) + ylim(0, 115) + 
  labs(title = "Union Rates and Party Control of Government", subtitle = "with ggplot2",
       x = "Left Control of Government", y = "Union Labor Force Percentage") + 
  theme_bw() + theme(legend.position = "none")
```

2. We will use Lowery et al.’s data on lobbying, the data ﬁle name is `constructionData.tab`. Lowery et al. is interested in testing whether the number of health establishments in a state aﬀects the participation rate of the health finance industry. Using their dataset, create a scatterplot either by using base R graphics or ggplot2 \footnote{Adapted from Monogan III, Political Analysis Using R}

```{r}
cons <- read.delim("day3data/constructionData.tab")
```

```{r}
head(cons)
```

```{r}
cons_lm <- lm(partratebusness ~ supplybusiness, cons)
summary(cons_lm)
```

```{r}
plot(x = cons$supplybusiness, y = cons$partratebusness, pch = 19, cex.main = 1, cex = 0.8, 
main = "Health Establishments and Lobby Participation Rate",
     ylab="Lobby Participation Rate", xlab="Number of Health Establishments")
abline(cons_lm, lty = 1, col = "blue") 
```

\noindent \textit{Note: Some of the material and practice exercises in this handout are adapted from Kosuke Imai's and James Monogan's textbooks and teaching materials.} 
