---
title: 'Introduction to R, Day 1 Handout'
output: pdf_document 
#bibliography: references.bib
urlcolor: red 
---
\setcounter{table}{0}
\renewcommand{\thetable}{A\arabic{table}}
\renewcommand{\thefigure}{A\arabic{figure}}

```{r setup, include=FALSE}
library(knitr)
library(formatR)
knitr::opts_chunk$set(echo = TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=55))
```

In this handout, we cover the following topics and R commands. The topics covered corresponds to Kosuke Imai's QSS Chapter 1. 

## Topics 
- Organizing your files
- Quick Recap: RStudio Interface to R
- Using R as a Calculator 
- Preparing and Working in an R Script for Analysis 
- Creating and Using Objects in R
- Working with Vectors
- Basic Operations on Vectors / Basic Functions
- Reading / Loading Data into R
- Working with Data Frames: Summarizing Data
- Packages
- Environment / Workspace
- Saving Objects
- Creating Data Files 
- Help Files 
- The Structure of R 

## R Commands 
- Doing arithmetic operations `+`, `-`, `*`, `^`, `sqrt()`, `log()` 
- Setting and displaying the working directory using `setwd()` and `getwd()`
- Creating an object using the assignment operator `<-` 
- Using `class` to obtain information about types of objects 
- Using `c()` to create and combine vectors 
- Using `[]` to access elements of vectors 
- Basic functions/basic operations on vectors: `length()`, `min()`, `max()`, `range()`, `mean()`, `median()`, `sum()`, `abs()`, `unique()`
- Using `names()` to assign or access names of objects
- Reading/loading data using `read.csv()`, `read.table`, and `load()`
- Summarizing the data using `head()` and `summary()`
- Obtaining the dimension, number of rows and columns, of data with `dim()`, `nrow()`, and `ncol()`
- Accessing variables from data frames with `$` 
- Creating new variables with `$`
- Using `[row, column]` to recover elements of a data frame
- Summarizing variable categories with `table()` 
- Listing objects in the workspace with `ls()` 
- Saving objects with `save()` and the workspace with `save.image()` 
- Creating data ﬁles using `write.table()` and `write.csv()` 
- Obtaining help for R with `help()`, `?`, and `??` 

\clearpage

# Before We Start: Organizing Your Files

- Where do you want store all your work on this workshop? This will be important when we start working with R scripts and data sets. 

- I will recommend putting all the material in the same folder, and store it somewhere that you can easily find. For example, I created a folder named `RWorkshop2021' on my desktop. 

- We will turn back to this point later when we discuss setting our working directory, which is the place where we want R to look for our files. 

# Using R as a Calculator

- First, we'll learn how to use R as a calculator. 

- We will use the console (the bottom left window) where we can directly enter R commands. You can think console like your scratch paper. Here, you can do some quick computations or try out some codes. 

- Some of the key arithmetic commands are: addition `+`, subtraction `-`, multiplication `*`, division `/`, exponentiation `^`, taking the square root of a number `sqrt()`, taking the log of a number `log()`. Parentheses `( )` specify the order of operations.

- Type the following arithmetic operations in your console and and hit \<enter\> or \<return\>. You will see that the console show the output of the commands you type.

```{r}
13 + 2
```

```{r}
13 - 2
```

```{r}
13 * 2
```

```{r}
13 / 2
```

```{r}
13 ^ 2
```

```{r}
sqrt(25)
```

```{r}
log(15)
```

- R follows the rules of the order of operations (PEMDAS), so we use parentheses `()` to set the order of operations. 
```{r}
13 * (11 - 2)
```

### $\spadesuit$ Coding Tip

\noindent R ignores spaces, so `5+3` will return the same result as `5 + 3`. However, we add a space before and after the arithmetic operators to make it easier to read.

# Preparing and Working in an R Script for Analysis 

- When you want to start working on a new analysis in R, you can follow a \textit{four step procedure:} 
\begin{enumerate}
\item Open an R Script
\item Set your working directory
\item Save your R script
\item Start working inside your R script 
\end{enumerate}

## Opening an R Script 

- If you haven't already, you can open an R Script by clicking:

**File $\rightarrow$ New File $\rightarrow$ R Script **

- Text editor window (the the upper left window) is the main place where we will execute R commands in our R script. 

## Setting working directory with `setwd()'

- When we use R with RStudio, the first thing we always want to do is to set our working directory. In other words, we want to tell R a folder in our computer where we will save your R scripts and datasets.

- Earlier, we created a folder for the workshop $-$ `RWorkshop2021' $-$ with this in mind. 

- Inside this folder, we can also create specific folders for days of the workshop. For example, I created a folder called `RWDay1' for today's workshop. I have also put all my relevants datasets to this folder.

- I suggest creating two folders inside your main workshop folder. The first folder can be a general folder for today. Once you create this folder, you can also create a dataset folder for today as well. 

- Now, we are ready to set our working directory. The easiest way to do this is to do it from the Session menu. So, let's select: 

**Session $\rightarrow$ Set Working Directory $\rightarrow$ Choose Directory...**

- This will tell R the folder that you will use throughout your analysis. When you’re done, you’ll notice some changes. These changes will give you hints to check whether you were able to set your working directory. 

- First, you’ll notice that the address in grey at the top of your console window has changed to show that you’re in the folder that you selected for Day 1. You can tell this by looking at the path above the console. 

Mine looks like this:

\begin{figure} [h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=.8\linewidth]{console.png}\\
\end{tabular}
\end{center}
\end{figure}

- Second, you will also see a change in your bottom-right window, files tab After setting your directory, update the ﬁle view and you’ll see that your working directory has changed to the folder you selected. Here, you can see what's inside your folder. 

Mine looks like this:

\begin{figure} [h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=.8\linewidth]{files2.png}\\
\end{tabular}
\end{center}
\end{figure}

### $\odot$ Note on Working Directory 

`Working directory` is an important element in the structure of R because it’s where R looks to ﬁnd ﬁles. It’s where our files land when we save them. In the future, you might ﬁnd yourself wondering ``Why can’t R ﬁnd this ﬁle? I can see it here!`` Usually the reason is that although the ﬁle is there, the working directory is somewhere else And, where you look and where R look is different. To avoid this confusion, it’s a good  practice to keep the ﬁles that you can see in the bottom right window in sync with your working directory. You can always do that by pressing the grey arrow at the top (next to the path).

- Suppose that you forget or not sure where you had set your working directory...When you type `getwd()` in your console, R will tell you where you had set your current working directory. 

```{r, echo=T}
getwd()
```

- What if you want to change your working directory? You can follow the same procedure: **Session $\rightarrow$ Set Working Directory $\rightarrow$ Choose Directory...**. You also have another option. You can use the `setwd()` and manually change your desired working directory.

```{r}
setwd("/Users/burcukolcak/Desktop/RWorkshop2021/RWDay1")
```

## Saving an R Script

- Now, we are ready to save our R script to our working directory, which corresponds to our main folder. To save your script: 

**File $\rightarrow$ Save As** 

- Make sure that the folder you save your R script is the same folder you have choosen for your working directory. 

- All of our R scripts will have the `.R extension'.

## Preparing an R Script

- Now, before we explore new topics, we want to prepare our R script. 

- The comment character `#` will play an important role when we work in an R script. This will be especially important as our code gets more complex. It allows us to annotate our R code behind commands or on a separate line. Everything that follows `#` will be ignored by R. We should annotate our own code for two main reasons: for us to remember what each part of the code does, and for others to understand what we are trying to do in our analysis. 

### $\spadesuit$ Coding Tip 

It is common to use a double comment character `##` if a comment is made for a seperate line and use a single comment character `#` if a comment is made within a line after an R command. Below is an example: 

```{r,results='hide'}
## Double comment vs. single comment

## Commenting on a seperate line 
5 + 3 

5 + 3 # commenting behind a command 
```

- Now, we are ready to prepare our first R script.

- First things first, let's give a title to our R script using the double comment character `##`. 

- Earlier, we discussed that we always want to set our working directory first. Let's set our your working directory and display our working directory. And, annotate each command in seperate lines. 

```{r, echo=T,results='hide'}
####################

## R Workshop Day 1

## set working directory 
setwd("/Users/burcukolcak/Desktop/RWorkshop2021/RWDay1")

## check working directory 
getwd()
```

- Next, let's do some basic arithmetic operations similar to what we did before. This time we will execute these commands in our R script. 

- Type the following arithmetic operations in your R script and use the symbol `#` comment character for explaining what you do: 

```{r, echo=T}
## Some Basic Arithmetic Operations 

## Add 3 + 7 
3 + 7

13 - 3 # subtract 13 - 3

## Take the square root of 100 using sqrt()
sqrt(100)
```

In the end, we want to see something like follows: 
\begin{figure} [h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=.6\linewidth]{prescript2.png}\\
\end{tabular}
\end{center}
\end{figure}

## Executing Commands in our R script

\begin{itemize}
\item To execute commands on your R script: 
\begin{itemize}
\item Place your cursor at the end of the line of code (far right), and hit <command + return> on a Mac; <control + return> on Windows. 
\item Or, place your cursor at the end of the line of code (far right), and click `Run' in the top right corner of the upper-left window.
\end{itemize}
- After you execute each command, the result wil pop up in your console. \footnote{Explanation of this part taken from Prof.Katie McCabe's MLE book.}
\end{itemize} 

# Creating and Using Objects

- In R, we can store various information as an object with a name of your choice. Objects are very handy because they serve as ``shortcuts'' to some piece of information or data that we would like to save and use throughout our analysis. 

- We use the assignment operator `<-` to assign a name to an object. 

### $\spadesuit$ Coding Tip 

There are some specific rules we need to follow when we choose a name for our objects in R. 
1. Object names cannot begin with numbers (but it can contain numbers) and no spacing is allowed. 
2. In general, we should avoid special characters such as \% and \$ because these have specific meanings in R. 
3. We should also try using intuitive and informative names for our objects. 
4. Object names are case sensitive. For example, if your object name is `butterfly`, you will get an error when you type `Butterfly` to print the value of the object. 

- In RStudio, once you create an object, you will see the object you created in the your Workspace/Environment (the upper right window). 

- Previously, we added `3 + 7`. Let's store the result of this calculation as an object named sum37. 

```{r}
## Create an object called sum 37
sum37 <- 3 + 7 
```

- By default, R will print the value of the object to the console if we just type the name of the object and run this command. 

```{r}
## display the value of sum37 
sum37 
```

- Or, you can explicitly tell R to print the value of the object by using the `print()` function.

```{r}  
## print the value of sum37 
print(sum37)
```

- Let's create another object to store information about a previous calculation we did `13-3`. Let's name this object `result`. 

```{r}
## Create an object called result 
result <- 13 - 3 
## display the value of result 
result 
```

- Let's use the same object name for storing information about another calculation. 

```{r}
## Create an object called result 
result <- 5 * 12 
## display the value of result 
result 
```

- Oops..We lost the previous value we assigned to this object! If we assign a different value to the same object name, then R will change the value of the object we previously created. Therefore, we need to be careful about this if we plan to use the previously assigned information later. To avoid this, for example, we can instead name the first object `result` and second object `result2`

```{r}
## Create an object called result 
result <- 13 - 3 
result 
## Create an object called result2
result2 <- 5 * 12 
result2
```

- Previously, we noted that object names are case sensitive. Let's see what happens when we write `Result` instead of `result`. 

- We can also store information about other types of objects different from numbers. For example, we can store information about characters as follows. In character strings, spacing is allowed. 

```{r}
## Create an object called statename 
statename <- "New Jersey"
statename
```

- How does R recognize diffrent types of objects? In other words, how does it differentiate between a number and character strings? R recognizes different types of objects by assigning each object to a `class`. Classes are important in R because it allows R to categorize objects into classes and then consider this when performing appropriate operations. 

- When we want to learn the class of an object, we will use the  function `class()`.  

-  A number is stored as a `numeric` object whereas a character string is recognized as a `character` object. 

```{r}
class(result) # numeric variable 
class(statename) # chracter variable 
class(log) # function 
```

- There are many other classes in R, and we will revisit this topic throughout the workshop. 

# $\mho$ Practice Exercises

U.S. Treasury Securities held by foreign countries increased from 3708.8 billion in January of 2010
to 4009.2 billion in June of 2010. In January, China held 889.0 billion in U.S. Treasury Securities.
By the close of June, Chinese holdings decreased to 843.7 billion. In contrast, Japanese holdings
increased from 765.4 billion in January to 803.6 billion in June. \footnote{Adapted from Professor Kosuke Imai's teaching materials}

1. Open a new R script and give it the following title `Exercise1`. 

2. Copy this question to your script and annotate it using the `##`  comment character.

3. Create two seperate objects called `us01` and `us06` to store information about holdings of the US in January of 2010 and June of 2010 by. Print the values of these objects by typing their names. Annotate your steps using the `##` character. 

```{r echo=F,results='hide'}
## holding of the US in January 
us01 <- 3708.8
us01
## holding of the US in June  
us06 <- 4009.2
us06
```

4. Store information about holdings of the China in January of 2010 and June of 2010 by creating two seperate objects called `china01` and `china06`. Print the values of these objects by typing their names. Annotate your steps using the `##` character. 

```{r echo=F,results='hide'}
## holding of the China in January 
china01 <- 889.0
china01
## holding of the China in June 
china06 <- 843.7
china06 
```

5. Store information about holdings of the Japan in January of 2010 and June of 2010 by creating two seperate objects called `japan01` and `japan06`. Print the values of these objects by typing their names. Annotate your steps using the `##` character. 

```{r echo=F,results='hide'}
## holding of the Japan in January 
japan01 <- 765.4
japan01
## holding of the Japan in June 
japan06 <- 803.6
japan06
```

6. Create three new objects for storing information about the differences between holdings of January of 2010 and June of 2010 for each country. You can call these objects `diffus`, `diffchina`, and `diffjapan` respectively. Print the values of these objects by typing their names. Annotate your steps using the `##` character. 

```{r echo=F,results='hide'}
## Difference of the US holdings 
diffus <- us06 - us01 
diffus
## Difference of the China holdings 
diffchina <- china06 - china01
diffchina
## Difference of the Japan holdings 
diffjapan <- japan06 - japan01
diffjapan
```

7. Check the class of objects `diffus`, `diffchina`, and `diffjapan` using the `class()` function. 

```{r echo=F,results='hide'}
class(diffus)
class(diffchina)
class(diffjapan)
```

\clearpage

# Working with Vectors 

- A vector is simply a collection of information (numbers or character strings) stored in a specific order. 

- In R, vectors are constructed by using the `c()` function, which is used to *combine* 
multiple values together into a vector.

- This will allow us to  create a `data vector` containing multiple values with commas separating different elements of the vector we are creating. We will see that vectors would become very useful for many operations we do in R. 

## Creating Vectors 

- To understand the logic of vectors, we are going to create a set of vectors, for some columns of  the following table. This table contains data from V-Dem about the top 5 main autocratizing countries between 2009-2019 \footnote{Source: V-Dem Democracy Report 2020}

\begin{figure} [h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=.8\linewidth]{vdem.png}\\
\end{tabular}
\end{center}
\end{figure}

- Let's create vectors for some columns by using the `c()` function. We enter the data one value at a time, each separated by a comma.
```{r}
## Create a vector the change variable 
change <- c(-0.36, -0.36, -0.33, -0.27, -0.25)
change
## Create a vector the ldi 2009 variable 
ldi2009 <- c(0.76, 0.46, 0.83, 0.53, 0.76)
ldi2009 
## Create a vector the ldi 2019 variable 
ldi2019 <- c(0.40, 0.10, 0.50, 0.25, 0.51)
ldi2019
## Create a vector the regime 2009 variable 
regime2009 <- c("liberal democracy", "electoral democracy", "liberal democracy", "liberal democracy", "electoral democracy") 
regime2009
```

# $\mho$ A Short Practice Exercise 

Now, its your turn. 

1. Create a vector for the column regime 2019

```{r echo=F,results='hide'}
regime2019 <- c("electoral autocracy", "electoral autocracy",  "electoral democracy", "electoral autocracy", "electoral democracy") 
regime2019
```

2. What is the class of this vector? 

```{r echo=F,results='hide'}
class(regime2019) 
```

- We can subtract vectors from each other. By subtracting `ldi2019` from `ldi2009`, we can calculate the change column. Let's call this change2 

```{r}
## calculate the diff between ldi 2019 and ldi 2009 
change2 <- ldi2019 - ldi2009 
change2
```

- We can also use the `c()` function to combine multiple vectors 

```{r}
## combine ldi2009 and ldi 2019 
ldi.all <- c(ldi2009, ldi2019) 
ldi.all
```

## Accessing Elements of a Vector / Indexing 

- What if we want to access specific elements of a vector? We will use the the square brackets, 
`[ ]` to access specific elements within the vector. This is called `indexing'. 

- For example, suppose that we want the 3rd value of the vector we created for `change`. This corresponds to the country change in Poland's democracy score. 
```{r}
change[3] # access and print the 3rd value of the vector change  
```

- Multiple elements can be extracted via a vector of indices within square brackets. 
```{r}
change[c(3, 4)] # access and print the 3rd and 4th value of the vector change
```

- Also within square brackets the dash symbol `-`, removes the corresponding element from a vector. Note that none of these operations change the original vector.
```{r}
change[-3] # remove the third value of the vector change 
```

# Basic Operations on Vectors/Basic Functions 

- One way we will use R a lot is through functions, which will allow us to act on or get information about vectors and other objects. 

- The following functions will be useful for doing basic operations on vectors:

- `length(x)` calculates the number of elements/entries in the `x` vector.
- `min(x)` calculates the smallest value in the `x` vector.
- `max(x)` calculates the largest value in the `x` vector .
- `mean(x)` calculates the average value in the `x` vector (that is the sum of the entries divided by the number of entries).
- `sum(x)` calculates the sum of the values in the `x` vector.
- `abs()` returns the absolute values of the `x` vector. 
- `unique(x)` returns a set of unique values from the `x` vector. 

- Turning back to our running example, let's use these functions to get information on our change vector. 

```{r}
length(change)
min(change)
max(change)
mean(change)
sum(change)
abs(change)
unique(change)
```

## Giving Names to Vectors 

- We can also give names to our vectors. Recall the change vector that we have been working on. 

```{r}
change
```

- Although we know that each element of this vector represents a country, it is difficult to remember these. So, we might want to assign names to elements of this vector. 

- Currently no names assigned to the change vector yet. 
```{r}
names(change) 
```

- The `names()` function can access and assign names to elements of a vector. 
```{r}
names(change) <- c("Hungary", "Turkey", "Poland", "Serbia", "Brazil")
change
```

### $\spadesuit$ Coding Tip  

\noindent Systematic spacing and indentation are essential when we are coding. For example, we added spaces after a comma. We also place spaces around all binary operators such as `<-`, `=`, `+`, and `-`. 

# Reading/Loading Data into R

- We can slowly start working with a real dataset to obtain basic information about it. 

- But, first, we need to know how to read data into R. 

- We will use different versions of the `UNpop` dataset. On Slack, you will find three different versions of the same dataset `UNpop.csv`, `UNpop.tsv` and `UNpop.R`. After downloading these datasets, we need put them into our dataset folder. 

- Two functions can be particularly useful when we first load a dataset into R:
1. We can use the `head()` to check first five rows of this dataset. 
2. We can use the `summary()` function to summarize the variables of this dataset. 

These two functions will help us check whether we have loaded in the data without error. 

Now, we will load each of these data files respectively. 

### Reading in a comma-separated `.csv` file

```{r}
UNpop <- read.csv("day1data/UNpop.csv")
```

```{r}
head(UNpop)
```

```{r}
summary(UNpop)
```

### Reading in a tab-delimited .tsv or .txt ﬁle

```{r}
UNpop2 <- read.table("day1data/UNpop.tsv", header = TRUE)
```

```{r}
head(UNpop2)
```

```{r}
summary(UNpop2)
```

- For the .tsv file, we set the parameter header to TRUE to tell R that the ﬁrst row of the ﬁle should be used to name each column. Let's see happens if we don't use this parameter: 

```{r}
UNpop3 <- read.table("day1data/UNpop.tsv")
```

```{r}
head(UNpop3)
```

### Reading in an R data file

```{r}
load("day1data/States.Rdata")
```

```{r}
head(States)
```

```{r}
summary(States)
```

# Working with a Data Frame: Summarizing Data

- When we load datasets into R, R will store them as `data frame` objects. 
```{r}
class(UNpop)
```

- A `data.frame` is an object in R that is basically like a spreadsheet which contains one column for each variable, and one row for each observation.   

- Below are some basic operations we can do on data frames to obtain information about the UNpop data set: 

- `names()` returns the column (variable) names of the data.

```{r}
names(UNpop)
```

- `ncol()` returns the number of columns in the data.
```{r}
ncol(UNpop)
```

- `nrow()` returns the number of rows of the data.
```{r}
nrow(UNpop)
```

- `dim()` returns a vector of the number of rows and the number of columns (the dimension of the data).
```{r}
dim(UNpop)
```

- `summary()` provides a summary of each variable in the data.
```{r}
summary(UNpop)
```

## Subsetting a Data Frame I 

- We'll often need to access different parts of a data frame. For example, we will want to extract information about a specific variable and find its mean. Or we will want to extract the 5th observation of a specific variable. This corresponds to subsetting a data frame in R. 

- To select a particular variable from the data frame, we can use the dollar `$` operator.

- In the UNpop data frame, to select the year variable we will use the `$` operator. Here,  `UNpop$year` will be a vector of just the `year` column of the `UNpop` data frame.
```{r}
UNpop$year
```

## Creating A New Variable with `$` 

- We will also use the `$` sign to create a new variable. Suppose that we want to create a new variable that subtracts 5 from each year. 

```{r}
UNpop$year2 <- UNpop$year - 5
UNpop$year2
```

# $\mho$ A Short Practice Exercise

1. Use the `$` to select the variable `world.pop` variable from the `UNpop` data frame. 

```{r echo=F,results='hide'}
UNpop$world.pop
```

2. Calculate the mean world population over this time period

```{r echo=F,results='hide'}
mean(UNpop$world.pop) 
```

## Subsetting a Data Frame II 

- You can also use the square brackets `[ ]` to subset from the data frame.

- As with vectors, square brackets may be used to call specific portions of the data frame. As opposed to vectors, however, with a data frame we need to tell R whether we want to subset the rows or columns. 

- To do so, we will use a comma and the bracket will have the following form: `[rows, columns]` where the expression before the comma will select the rows and the expression after the comma will select the columns. This will allow us to call specific rows and columns by either row/column numbers or row/column names

- `UNpop[,"world.pop"]` will select the `world.pop` variable from `UNpop`: 
```{r}
UNpop[,"world.pop"]
```

- `UNpop[1,]` will select the first row of `UNpop`
```{r}
UNpop[1,]
```

- `UNpop[c(1,2,3),]` will select the first three rows of `UNpop`

```{r}
UNpop[c(1,2,3),]
```

- `UNpop[1:3, "world.pop"]` will select the first three values of the `world.pop` variable of `UNpop`

```{r}
UNpop[1:3, "world.pop"]
```

- Most of the time we will want to summarize a variable's categories. To do so, we can use the `table()` function. 

- For example, suppose that we want to summarize the values of the year variable in the UNpop data frame: 

```{r}
table(UNpop$year)
```

# Packages

- One of R’s mains strengths is the community of R users who contribute to the development of R with packages. 

- These packages are available through the Comprehensive R Archive Network

- Thus far, we haven't used any packages. We were able to do everything on `base R` functions. In other words, we used functions that already exist in R. 

- For using other functions,  we will install packages. Throughout the workshop, we will use various packages. For example, we will install a package called ``ggplot2` for data visualization in the following days. 

- Suppose that we wish to load a data file produced by another statistical software such as Stata with the .dta extension. R has a package called `readstata13` which is useful when dealing with Stata data files. 

- To use any package in R, if we are using that package for \textit{the first time},  we need to do two things:  

1. Install the package using `install.packages("packagename")`
2. Load the package using `library(packagename)`

### $\odot$ Note on Packages
 
\noindent Package installation needs only to occur once. This means that when you want to use the `readstate13` package next time, you only need to load the `library()` in the beginning of your R script. 

- Now, let's try to read a Stata file into R: 

```{r}
library(readstata13) # load package
UNpopdta <- read.dta13("day1data/UNpop.dta") #a STATA data file with .dta extension 
head(UNpopdta)
summary(UNpopdta)
```

# Environment/Workspace

- All the objects we created thus far appear in our **workspace/environment** (the upper right window). 

- If we want to list the set of all objects in our current R session. We can use the function `ls()`:

```{r}
ls()
```

The workspace will include all objects you created - variables, vectors, data frames, 
and a history of all commands you've run in this session.

# Saving Objects  

- The objects we create during an R session are only temporarily stored to
the workspace. Unless you save objects from the workspace before exiting R, you will lose all objects
and changes to data.

- To save any object, we can use `save(xxx, file = "yyy.RData")` function where `xxx` is the
object name and `yyy.RData` is the file name. Multiple objects can be listed and they will be stored as a `single .RData file`. The extension `.RData` should always be used for the file name.
Unless the full path is specified, objects will be saved to the working directory.

- For example, suppose we want to save one of the objects we created  before:  
```{r}
save(change, file = "change.RData")
```
 
- To save the entire workspace (all objects), we can use the `save.image()` function with a `.RData`
file name.

```{r}
save.image(file = "Handout1.RData")
```

- To access the objects saved in the .RData file, simple double click the file or use the load()
function as above

### $\odot$ Note on Saving Workspace 

\noindent When you close RStudio, you will be asked if you want to save your workspace. In general, we will say *no* to this question because we have all the ingredients in our script to replicate our analysis again. We usually save scripts, not the work space. 

# Creating Data Files 

- If you create new variables by building on an existing dataset, you may also want to generate a text or csv ﬁles. 

- You can use `write.table()` or `write.csv()` to generate data files. 

- Suppose that we want to generate new data files of the UNpop data frame.  
```{r}
write.csv(UNpop, file = "UNpopnew.csv") # generate a csv file 
write.table(UNpop, file = "UNpoplast.csv") # generate a text file 
```

# Help Files

- If you want to access RStudio’s help functions, click on the `Help` tab in the bottom left window, and type in your question.

- If you want to access help ﬁles from the text editor window (command line), type in `?command`, e.g. `?table` to learn about the table function. 

- If you do not remember the exact name of an R command, type `??tab` and R will use `fuzzy matching` to suggest some commands.

# The Structure of R

So far, we have learned the following elements of R:

- The **R script**.  This is where you conduct your main analysis. We use to the text editor window (the upper left window) to work inside an R Script. Your script contains all of the R commands necessary to replicate your analysis, as well as comments explaining the code.

- The **R console**. If we just want to see what a command does, or quickly calculate some quantity, we can enter your command directly into the R console.

- The **working directory**.  This is the directory from where R will find your data files. The first thing we do when we start an analysis is to set and display our working directory to be sure that we and R are at the same page :) 

- The **environment/workspace**. Workspace contains set of all objects in your current R session. 

\clearpage 

# $\mho$ Practice Exercises

**Dataset: turnout.csv (in day1data folder)**

Surveys are frequently used to measure political behavior such as voter turnout, but some researchers are concerned about the accuracy of self-reports. In particular, they worry about possible \textit{social desirability bias} where, in postelection surveys, respondents who did not vote in an election lie about not having voted because they may feel that they should have voted. Is such a bias present in the American National Election Studies (ANES)? ANES is a nationwide survey that has been conducted for every election since 1948. ANES is based on face-to-face interviews with a nationally representative sample of adults \footnote{Adapted from Kosuke Imai's QSS book, Chapter 1}

Table below displays the names and descriptions of variables in this data set: (CORRECT THIS TABLE!)

-----------------------------------------------------------------------------
 Name                    Description
 ----------------------- ----------------------------------------------------
`year`                   election year 
 
`ANES`                   ANES estimated turnout rate
 
`VEP`                    voting eligible population (in thousands)

`VAP`                    voting age population (in thousands)
                      
`total`                  total ballots cast for highest ofﬁce (in thousands)
 
`felons`                 total ineligible felons (in thousands)
 
`noncitizens`            total noncitizens (in thousands)
 
`overseas`               total eligible overseas voters (in thousands)
 
`osvoters`               total ballots counted by overseas voters (in thousands)
-----------------------------------------------------------------------------

1. Load the dataset into a data frame called `turnout`. 

```{r echo=F,results='hide'}
turnout <- read.csv("day1data/turnout.csv")
```

2. Check the first six rows of this data frame. 

```{r echo=F,results='hide'}
head(turnout)
```

3. Obtain a summary of the data. What is the range of years covered in this dataset?

```{r echo=F,results='hide'}
summary(turnout)
```

3. Check the dimensions of the data. How many rows and columns are there?

```{r echo=F,results='hide'}
dim(turnout)
nrow(turnout)
ncol(turnout)
```

4. Calculate the turnout rate based on VAP (the voting age population). Note that for this dataset, we must add the total number of eligible overseas voters since the VAP variable does not include these individuals in the count. Add years as informative labels to see the turnout change over time. 

```{r echo=F,results='hide'}
## Calculate VAP-based turnout
vaptr <- turnout$total / (turnout$VAP + turnout$overseas) * 100 
vaptr
## Add informative labels using names()
names(vaptr) <- turnout$year 
vaptr
```

5. Calculate the turnout rate based on VEP (the voting eligible population). Add years as informative labels to see the turnout change over time. 

```{r echo=F,results='hide'}
## Calculate VEP based turnout
veptr <- turnout$total / turnout$VEP * 100
veptr
## Add informative labels using names()
names(veptr) <- turnout$year
veptr
```

6. Calculate the difference between VEP and VAP based turnouts. 

```{r echo=F,results='hide'}
## Calculate the difference between VEP and VAP based turnout
diff <- veptr - vaptr 
names(diff) <- turnout$year 
diff
```

7. Compute the differences between the VAP and ANES estimates of turnout rate. How big is the difference on average? What is the range of the differences?

```{r echo=F,results='hide'}
## Comparison between VAP and ANES
diff2 <- turnout$ANES - vaptr
## You can use summary for getting info on mean and range 
summary(diff2)
## Or, you can use mean and range functions
mean(diff2)
range(diff2)
```

8. Conduct the same comparison for the VEP and ANES estimates of voter turnout.

```{r echo=F,results='hide'}
## Comparison between VAP and ANES
diff3 <- turnout$ANES - veptr
## You can use summary for getting info on mean and range 
summary(diff3)
## Or, you can use mean and range functions
mean(diff3)
range(diff3)
```

9. (optional) Calculate the VEP turnout rate for presidential elections and midterm elections. Note that the data set excludes the year 2006. Does the bias of the ANES estimates vary across election types?

```{r echo=F,results='hide'}
## Check the elections in the data
turnout$year
## Presidential elections: odd entries of vector (1st, 3rd…) plus the last
pres <- c(1, 3, 5, 7, 9, 11, 13, 14) 
mids <- c(2, 4, 6, 8, 10, 12)
## turnout during presidential elections, subsetting vector 
turnout$year[pres]
turnout$year[c(1, 3, 5, 7, 9, 11, 13, 14)] # this yields the same result 
## turnout during midterm elections, subsetting vector 
turnout$year[mids]
## turnout during presidential elections 
pveptr <- veptr[pres] 
names(pveptr) <- turnout$year[pres] 
pveptr
## turnout during midterm elections 
mveptr <- veptr[mids] 
names(mveptr) <- turnout$year[mids] 
mveptr
## average difference between presidential and midterm for VEP based turnout
mean(pveptr) - mean(mveptr)
```

10. (practice for later) Calculate the ANES turnout rate for presidential elections and midterm elections. Note that the data set excludes the year 2006. Does the bias of the ANES estimates vary across election types?

\noindent \textit{Note: Some of the material and practice exercises in this handout are adapted from Kosuke Imai's teaching materials. His materials are publicly avaliable and provide great resources for learning and practicing R.}
